### 2024-06-28 paper/tex/02-overview.tex
edit

Observe the consistent polarization pattern of types. Consider the type for
$[[double]]$, which in \systemf would be $[[Int → Int]]$. In polarized form, it
becomes $[[↓(Int → ↑Int)]]$---a \emph{suspended} function (indicated by the
downshift) that accepts an integer $[[Int]]$ and yields an integer
\emph{computation} (indicated by the upshift) $[[↑Int]]$.

@@

\begin{itemize}
\item [$\pm$] $[[na]]$ and $[[pa]]$
  denote negative and positive type variables, respectively.
  The variables are either introduced by the corresponding quantifiers 
  ($[[∀]]$ and $[[∃]]$, respectively) or declared in the type context;
\item [$-$] a function $[[iP → iN]]$ takes a \emph{positive} input $[[iP]]$ and
returns a \emph{negative} output $[[iN]]$; the function type itself is
\emph{negative}, so the `arrow' operator is right-associative: 
$[[iP → iQ → iN]]$ is equivalent to $[[iP → (iQ → iN)]]$;
\item [$-$] similarly to functions, a polymorphic abstraction $[[∀pas.iN]]$
  quantifies a negative type $[[iN]]$ over a list of positive type variables $[[pas]]$;
\item [$+$] dually, $[[∃nas.iP]]$ binds negative variables in a positive type $[[iP]]$;
\item [$\pm$] an upshift $[[↑iP]]$ and a downshift $[[↓iN]]$ change the polarity of a type
  from positive to negative and vice versa.
  At the level of terms, the constructors for shift types are
  $[[return v]]$ and $[[{c}]]$ (thunked computation), respectively.
\end{itemize}

\paragraph*{Syntactic Conventions}
We always consider terms and types up to alpha-equivalence, and capture-avoiding
substitution. Besides, we assume the quantification is associative,
for example, $[[∀pas.∀pbs.iN]]$ and $[[∀pas,pbs.iN]]$
represent the same type.

@@

In \cref{fig:declarative-terms}, we define the language of terms of \fexists.
The language is a combination of \systemf and \CBPV constructs: the
terms are stratified into values and computations; there are return 
and thunk constructors, as well as several forms of let bindings to
sequentialize the computation/application with or without type annotation;

\begin{itemize}
    \item [$+$] $[[x]]$ denotes a term variable.
      Following the \CBPV stratification, we only have \emph{positive} (value)
      term variables;
    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
        or suspended computation;
    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
        positive and negative terms;
    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
        are standard lambda abstractions. Notice that we require
        the type annotation for the argument of $[[λ]]$;
    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
    \item [$-$] computational let form $[[ let x : iP = c; c']]$ 
      operates similarly to a monadic bind.
      It binds the result of a computation $[[c : ↑iP]]$ 
      to a variable $[[x : iP]]$, and continues with a computation $[[c']]$;
    \item [$-$] applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
        $[[let x = v ( args ) ; c]]$ can be viewed as a special case of the
        computational let, when the first computation is a function application
        $[[v]] ([[args]])$.
        They take a suspended computation $[[v]]$, apply it to a list of
        arguments, bind the (pure) result to a variable
        $[[x]]$, and continues with a computation $[[c]]$. If the resulting type
        of the application is unique, one can omit the type annotation, as in
        the second form: it will be inferred by the algorithm;
    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
        is the standard eliminator of an existential type (unpack):
        expecting $[[v]]$ to be of an existential type,
        it binds the packed negative types to a list of 
        variables $[[nas]]$, binds the body of the existential
        to $[[x]]$, and continues with a computation $[[c]]$.
\end{itemize}

@@

\paragraph*{Subtyping} 
The inference algorithm's complexity mainly lies in the subtyping relation,
which is, in particular, responsible for polymorphic instantiation. It's this
subtyping relation that enables us to apply polymorphic terms in situations
where their specific instantiations are anticipated.

@@

\paragraph*{Equivalence}
The system introduces another complexity: the equivalence induced by subtyping
($[[Γ ⊢ iN ≈ iM]] \defiff [[Γ ⊢ iN ≤ iM]] \text{ and } [[Γ ⊢ iM ≤ iN]]$) is not
straightforward. This relation extends beyond just alpha-equivalent types. As
the neighboring polymorphic quantifiers can be instantiated in arbitrary order.
the equivalence relation must permit the \emph{permutations} of the quantifiers.
For example, we have $[[· ⊢ ∀α⁺.∀β⁺.iN ≈ ∀β⁺.∀α⁺.iN]]$. Similarly, the
equivalence permits the \emph{removal/addition} of the unused quantifiers: $[[·
⊢ ∀α⁺.↑Int ≈ ↑Int]]$.

@@

In many type systems, the least upper bound boils down to the trivial syntactic equality.
Because of that, the set of constraints is typically resolved using \emph{unification}.
However, this problem is trickier in the presence of existential types. For example, 
in \fexists, the least upper bound of $[[↓↑Int]]$ and $[[↓↑Bool]]$ is $[[∃α⁻.↓α⁻]]$,
because the quantified $[[α⁻]]$ can be instantiated to either $[[↑Int]]$ or $[[↑Bool]]$.

To find this least upper bound, we use \emph{anti-unification}---the dual of
unification. While unification finds \emph{the most general instance} of two
given patterns, the anti-unification finds \emph{the most detailed} (under the
restrictions of the system) \emph{pattern} that matches two given types. In the
example above, the anti-unifier of $[[↓↑Int]]$ and $[[↓↑Bool]]$ is $[[↓α̂⁻]]$,
and in \fexists, this unifier is the most specific (in particular because
$[[↓↑α̂⁺]]$ is not considered as a candidate since only \emph{negative} variables are
existentially quantified).

>>>>

Observe the consistent pattern of the type polarization. Consider the type for
$[[double]]$---a function multiplying an input integer by two. In \systemf, it
would be $[[Int → Int]]$. In call-by-value polarized form, it becomes $[[↓(Int →
↑Int)]]$---a \emph{suspended} function (indicated by the downshift) that accepts
an integer $[[Int]]$ and yields an integer \emph{computation} (indicated by the
upshift) $[[↑Int]]$.

The type of $[[map]]$ is more complex. Similar to $[[double]]$, the
entire type is a suspended function, hence $[[↓]]$. The function takes two implicit type
arguments: $[[α⁺]]$ and $[[β⁺]]$, which indicated by the $[[∀]]$s.
The third argument of $[[map]]$---the applied function---has type $[[α⁺ → ↑β⁺]]$,
which is suspended by $[[↓]]$ since it is in a (positive) argument position. 
The last argument---$[[List α⁺]]$---represents a list of $[[α⁺]]$s and does not need
to be suspended as it is already a value. Finally, the positive return type,
$[[List β⁺]]$, is lifted to computations using $[[↑]]$.

The types of $[[len]]$---the length function for lists, $[[choose]]$---a
function that chooses between two input values, and $[[id]]$---the identity
function, are similarly polarized. A slightly more complex function $[[auto]]$
takes as an argument and returns something of the same identity type. However,
since the argument position is positive, the argument type is suspended by
$[[↓]]$.

The types of $[[len]]$ (the length function for lists), $[[choose]]$ (a function
that chooses between two input values), and $[[id]]$ (the identity function) are
polarized in a similar way. Another function, $[[auto]]$, takes an argument and returns
a function of the same identity type. However, since the argument position is
positive, the argument type is suspended using $[[↓]]$.

@@

\begin{itemize}
\item [$\pm$] $[[pa]]$ and $[[na]]$
  denote negative and positive type variables, respectively.
  The variables are either introduced by the corresponding quantifiers 
  ($[[∀]]$ and $[[∃]]$, respectively) or declared in the type context;
\item [$-$] a function $[[iP → iN]]$ takes a \emph{positive} input $[[iP]]$ and
returns a \emph{negative} output $[[iN]]$; the function type itself is
\emph{negative}, so the `arrow' operator is right-associative: 
$[[iP → iQ → iN]]$ is equivalent to $[[iP → (iQ → iN)]]$;
\item [$-$] similarly to functions, a polymorphic abstraction $[[∀pas.iN]]$
  quantifies a negative type $[[iN]]$ over a list of positive type variables $[[pas]]$;
\item [$+$] dually, $[[∃nas.iP]]$ binds negative variables in a positive type $[[iP]]$;
\item [$\pm$] an upshift $[[↑iP]]$ and a downshift $[[↓iN]]$ change the polarity of a type
  from positive to negative and vice versa.
  At the level of terms, the constructors for shift types are
  $[[return v]]$ and $[[{c}]]$ (thunked computation), respectively.
\end{itemize}

\paragraph*{Syntactic Conventions}
We always consider terms and types up to alpha-equivalence, 
and substitution is capture-avoiding. In addition, we assume the quantification is associative,
for example, $[[∀pas.∀pbs.iN]]$ and $[[∀pas,pbs.iN]]$
represent the same type.

@@

In \cref{fig:declarative-terms}, we define the language of terms of \fexists.
The language is a combination of \systemf and \CBPV constructs: the
terms are stratified into values and computations; there are return 
and thunk constructors, as well as several forms of let bindings
to sequence computations, bind function applications, 
and eliminate existentials, all with or without type annotations.

\begin{itemize}
    \item [$+$] $[[x]]$ denotes a term variable.
      Following the \CBPV stratification, we only have \emph{positive} (value)
      term variables;
    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
        or suspended computation;
    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
        positive and negative terms;
    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
        are standard lambda abstractions. Notice that we require
        the type annotation for the argument of $[[λ]]$;
    \item [$-$] $[[ let x = v ; c]]$ is a standard let form, binding
        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
    \item [$-$] computational let form $[[ let x : iP = c; c']]$ 
      operates similarly to a monadic bind.
      It binds the result of a computation $[[c : ↑iP]]$ 
      to a variable $[[x : iP]]$, and continues with a computation $[[c']]$;
    \item [$-$] applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
        $[[let x = v ( args ) ; c]]$ can be viewed as a special case of the
        computational let, when the first computation is a function application
        $[[v]] ([[args]])$.
        They take a function $[[v]]$, apply it to a list of
        arguments, bind the (pure) result to a variable
        $[[x]]$, and continues with a computation $[[c]]$. If the resulting type
        of the application is unique, one can omit the type annotation, as in
        the second form: it will be inferred by the algorithm;
    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
        is the standard eliminator of an existential type (unpack):
        expecting $[[v]]$ to be of an existential type,
        it binds the packed negative types to a list of 
        variables $[[nas]]$, binds the body of the existential
        to $[[x]]$, and continues with a computation $[[c]]$.
\end{itemize}

@@

\paragraph*{Subtyping} 
The inference algorithm's complexity mainly lies in the subtyping relation,
which is responsible for polymorphic instantiation. It is this
subtyping relation that enables us to apply polymorphic terms in situations
where their specific instantiations are anticipated.

@@

\paragraph*{Equivalence}
The system introduces another complexity: the equivalence induced by subtyping
($[[Γ ⊢ iN ≈ iM]] \defiff [[Γ ⊢ iN ≤ iM]] \text{ and } [[Γ ⊢ iM ≤ iN]]$) is not
straightforward. This relation extends beyond just alpha-equivalent types. As
the neighboring polymorphic quantifiers can be instantiated in arbitrary order,
the equivalence relation must permit the \emph{permutations} of the quantifiers.
For example, we have $[[· ⊢ ∀α⁺.∀β⁺.iN ≈ ∀β⁺.∀α⁺.iN]]$. Similarly, the
equivalence permits the \emph{removal/addition} of the unused quantifiers: $[[·
⊢ ∀α⁺.↑Int ≈ ↑Int]]$.

@@

In Damas Milner type inference, the least upper bound boils down to the trivial syntactic equality.
Because of that, the set of constraints is typically resolved using \emph{unification}.
However, this problem is trickier in the presence of existential types. For example, 
in \fexists, the least upper bound of $[[↓↑Int]]$ and $[[↓↑Bool]]$ is $[[∃α⁻.↓α⁻]]$,
because the quantified $[[α⁻]]$ can be instantiated to either $[[↑Int]]$ or $[[↑Bool]]$.

To find this least upper bound, we use \emph{anti-unification}---the dual of
unification. While unification finds \emph{the most general instance} of two
given patterns, the anti-unification finds \emph{the most detailed} (under the
restrictions of the system) \emph{pattern} that matches two given types. In the
example above, the anti-unifier of $[[↓↑Int]]$ and $[[↓↑Bool]]$ is $[[↓α̂⁻]]$,
and in \fexists, this anti-unifier is the most specific (in particular because
$[[↓↑α̂⁺]]$ is not considered as a candidate since only \emph{negative} variables are
existentially quantified).

### 2024-06-28 paper/tex/03-declarative-system.tex
edit

In this section, we present the declarative system of \fexists,
which serves as a specification of the type inference algorithm.
The declarative system is represented as a set of inference rules
and consists of two main subsystems: subtyping and type inference.
First, we present the declarative subtyping rules specifying 
when one type is a subtype of another; next, we discuss 
the equivalence relation induced by mutual subtyping;
finally, we present the type inference rules,
that refer to the subtyping and equivalence relation.
We conclude this section by discussing the relation between the
proposed type system and the standard \systemf.

@@

\paragraph*{Quantifiers}  
Symmetric rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} specify 
subtyping between top-level quantified types.
Usually, the polymorphic subtyping is represented by two rules
introducing quantifiers to the left and to the right-hand side of subtyping.
For conciseness of representation, we compose these rules into one.
First, our rule extends context $[[Γ]]$ with the quantified variables 
from the right-hand side ($[[pbs]]$ or $[[nbs]]$), 
as these variables must remain abstract.
Second, it verifies that the left-hand side quantifiers
($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively. 
This is the point of non-determinism as the instantiation can be chosen arbitrarily

The instantiation of quantifiers is modeled by substitution $[[σ]]$.
The notation $[[Γ2 ⊢ σ : Γ1]]$ specifies its domain and range.
For instance, $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
$[[σ]]$ maps the variables from $[[pas]]$ to (positive) types
well-formed in $[[Γ, pbs]]$.
This way, application $[[ [σ]iN ]]$ instantiates (replaces) every
$[[αi⁻]]$ in $[[iN]]$ with $[[σ]]([[αi⁻]])$.

\paragraph*{Invariant Shifts}
As mentioned above, an important restriction that we put on the subtyping system
is that subtyping of shifted types requires their equivalence, as shown in
\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}. The
reason for this is that if both of these rules were relaxed to the covariant
form, the subtyping relation would become equivalent to the standard subtyping
of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. However, it is
suggested after certain changes \ruleref{\ottdruleDOneShiftULabel} can be
relaxed to the covariant form, thereby increasing the expressiveness of the
system. These changes are discussed in \cref{sec:weakening-invariant}.

@@

\subsubsection{Properties of Declarative Subtyping}
\label{sec:decl-subtyping-properties}

@@

\begin{property}[Positive least upper bound exists]
  Suppose that $[[iP1]]$ and $[[iP2]]$ are positive types
  well-formed in $[[Γ]]$.
  Then there exists a least common supertype--- a type $[[iP]]$ such that
  \begin{itemize}
    \item $[[Γ ⊢ iP ≥ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$, and 
    \item for any $[[iQ]]$ such that $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$,
      $[[Γ ⊢ iQ ≥ iP]]$.
  \end{itemize}
\end{property}

@@

The normalization algorithm is shown in \cref{fig:type-nf}. The steps (i-ii) are
implemented by the \emph{ordering} function `$[[ord varset in iN]]$' and `$[[ord
varset in iP]]$'. For a set of variables $[[varset]]$, and a type, it returns a
list of variables from $[[varset]]$ that occur in the type in the \emph{order of
their first occurrence}. For brevity, we omit the formal definition of ordering
referring the reader to the appendix (\cref{appendix}).

@@

The declarative specification of the type inference is given in 
\cref{fig:declarative-inference}.
The positive typing judgment $[[Γ ; Φ ⊢ v : iP]]$ is read as 
``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the term $[[v]]$ is allowed to infer type $[[iP]]$'',
where $[[Φ]]$---the variable context---is defined standardly as
a set of pairs of the form $[[x : iP]]$. 
The negative typing judgment is read similarly.

The \emph{Application typing} judgment
infers the type of the application of a function to a list of arguments.
It has form $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$, 
which reads ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the application of a function of type $[[iN]]$ to the list of arguments $[[args]]$
is allowed to infer type $[[iM]]$''.

@@

\subparagraph*{Abstractions}
  The typing of lambda abstraction is standard. 
  Rule \ruleref{\ottdruleDTtLamLabel} first checks
  that the given type annotating the argument is well-formed,
  and then infers the type of the body in the extended context.
  As a result, it returns an arrow type of function from the
  annotated type of the argument to the type of the body.
  Rule \ruleref{\ottdruleDTTLamLabel} infers polymorphic $[[∀]]$-type. 
  It extends the type context with the quantifying variable $[[pa]]$ and 
  infers the type of the body. As a result, it returns a polymorphic type
  quantifying the abstracted variable $[[pa]]$ over the type of the body.

@@

\subparagraph*{Unpack}
  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
  First, it infers the normalized type of the existential package.
  The normalization is required to fix the order of the quantifying variables
  to bind them. Alternative approaches that do not require normalization
  will be discussed in \cref{sec:explicit-type-application}.
  After the bind, the rule infers the type of the body 
  and checks that it does not use the bound variables so that they do not 
  escape the scope.

@@

\paragraph*{Application of a polymorphic type $[[∀]]$}
  The complexity of the system lives in the rules whose output type is not
  immediately defined by their input and the output of their premises (\aka not
  mode-correct \cite{dunfield2020:bidirectional}). In our typing system,
  \ruleref{\ottdruleDTForallAppLabel} is such a rule: the instantiation of the
  quantifying variables is not known a priori. The algorithm we present in
  \cref{sec:algorithm} delays this instantiation until more information about it
  (in particular, typing constraints) is collected.

@@

The type polarization (\cref{fig:polarization}) is more involved, as there are
multiple ways to polarize a type: for instance, a variable $[[α]]$ can be
polarized to $[[α⁺]]$ or $[[α⁻]]$. The choice of polarization affects the
execution strategy of the program. 
In particular, a functional type can be represented in either positive (thunked, call-by-name) way:
$[[↓(iP → ↑iQ)]]$ or negative (call-by-value): $[[↓iN → iM]]$.
We chose the positive polarization: every \systemf type is translated to a
\emph{positive} type of \fexists. This choice makes it smoother to lift the
translation to the term level, which will be discussed next.

@@

The term-level translations between the systems are more complex. They are
defined as \emph{elaborations} of the typing derivations. In other words, the
input of the translation is not a term, but a whole typing derivation in the
corresponding system. For example, the subtyping elaboration $[[Γ ⊢ iN ≤ iM ⤳
ft]]$ constructs a \systemf function $[[ft]]$ of type $[[|iN| → |iM|]]$ that
witnesses the given \fexists-subtyping $[[Γ ⊢ iN ≤ iM]]$; a typing elaboration
$[[Γ; Φ ⊢ v : iP ⤳ ft]]$ builds a term $[[ft]]$---a \systemf counterpart of
$[[v]]$---of type $[[|iP|]]$.

>>>>

In this section, we present the declarative system of \fexists,
which serves as a specification of the type inference algorithm.
The declarative system is represented as a set of inference rules
and consists of two main subsystems: subtyping and type inference.
First, we present the declarative subtyping rules specifying 
when one type is a subtype of another. Next, we discuss 
the equivalence relation induced by mutual subtyping.
Finally, we present the type inference rules,
that refer to the subtyping and equivalence relation.
We conclude this section by discussing the relation between the
proposed type system and the standard \systemf.

@@

\paragraph*{Quantifiers}  
Symmetric rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} specify 
subtyping between top-level quantified types.
Usually, the polymorphic subtyping is represented by two rules
introducing quantifiers to the left and to the right-hand side of subtyping.
For conciseness of representation, we compose these rules into one.
First, our rule extends context $[[Γ]]$ with the quantified variables 
from the right-hand side ($[[pbs]]$ or $[[nbs]]$), 
as these variables must remain abstract.
Second, it verifies that the left-hand side quantifiers
($[[pas]]$ or $[[nas]]$) can be instantiated 
with an arbitrary substitution to continue subtyping recursively,
which introduces non-determinism.

The instantiation of quantifiers is modeled by a substitution $[[σ]]$.
The notation $[[Γ2 ⊢ σ : Γ1]]$ specifies its domain and range.
For instance, $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
$[[σ]]$ maps the variables from $[[pas]]$ to (positive) types
well-formed in $[[Γ, pbs]]$.
This way, application $[[ [σ]iN ]]$ instantiates (replaces) every
$[[αi⁻]]$ in $[[iN]]$ with $[[σ]]([[αi⁻]])$.

\paragraph*{Invariant Shifts}
As mentioned above, an important restriction that we put on the subtyping system
is that subtyping of shifted types requires their equivalence, as shown in
\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}. The
reason for this is that if both of these rules were relaxed to the covariant
form, the subtyping relation would become equivalent to the standard subtyping
of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. 
Relaxations of this condition are discussed in \cref{sec:weakening-invariant}.

% However, it is
% suggested after certain changes \ruleref{\ottdruleDOneShiftULabel} can be
% relaxed to the covariant form, thereby increasing the expressiveness of the
% system.  These changes are discussed in \cref{sec:weakening-invariant}.

@@

\subsection{Properties of Declarative Subtyping}
\label{sec:decl-subtyping-properties}

@@

\begin{property}[Positive least upper bound exists]
  Suppose that $[[iP1]]$ and $[[iP2]]$ are positive types
  well-formed in $[[Γ]]$.
  Then there exists a least common supertype---a type $[[iP]]$ such that
  \begin{itemize}
    \item $[[Γ ⊢ iP ≥ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$, and 
    \item for any $[[iQ]]$ such that $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$,
      $[[Γ ⊢ iQ ≥ iP]]$.
  \end{itemize}
\end{property}

@@

The normalization algorithm is shown in \cref{fig:type-nf}. The steps (i-ii) are
implemented by the \emph{ordering} function `$[[ord varset in iN]]$' and `$[[ord
varset in iP]]$'. For a set of variables $[[varset]]$ and a type, it returns a
list of variables from $[[varset]]$ that occur in the type in the \emph{order of
their first occurrence}. For brevity, we omit the formal definition of ordering
referring the reader to the appendix (\cref{appendix}).

@@

The declarative specification of the type inference is given in 
\cref{fig:declarative-inference}.
It consists of three main judgments: the positive typing judgment,
the negative typing judgment, and the application typing judgment.
The positive typing judgment $[[Γ ; Φ ⊢ v : iP]]$ is read as 
``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the term $[[v]]$ is allowed to infer type $[[iP]]$'',
where $[[Φ]]$---the variable context---is defined standardly as
a set of pairs of the form $[[x : iP]]$. 
The negative typing judgment is read similarly.
The \emph{Application typing} judgment
infers the type of the application of a function to a list of arguments.
It has form $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$, 
which reads ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the application of a function of type $[[iN]]$ to the list of arguments $[[args]]$
is allowed to infer type $[[iM]]$''.

@@

\subparagraph*{Abstractions}
  The typing of lambda abstraction is standard. 
  Rule \ruleref{\ottdruleDTtLamLabel} first checks
  that the given type $[[iP]]$ annotating the argument is well-formed,
  and then infers the type $[[iN]]$ of the body in the extended context.
  As a result, it returns an arrow type $[[iP → iN]]$.
  Rule \ruleref{\ottdruleDTTLamLabel} infers polymorphic $[[∀]]$-type. 
  It extends the type context with the quantifying variable $[[pa]]$ and 
  infers the type $[[iN]]$ of the body. As a result, it returns a polymorphic type
  $[[∀pa.iN]]$.

@@

\subparagraph*{Unpack}
  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
  First, it infers the normalized type of the existential package.
  The normalization is required to fix the order of the quantifying variables
  to bind them. Alternative approaches that do not require normalization
  will be discussed in \cref{sec:explicit-type-application}.
  After the bind, the rule infers the type of the body 
  and ensures that the bound variables do not escape the scope.

@@

\paragraph*{Application of a polymorphic type $[[∀]]$}
  The complexity of the system lives in the rules whose output type is not
  immediately defined by their input and the output of their premises (\aka not
  mode-correct \cite{dunfield2020:bidirectional}). In our typing system,
  \ruleref{\ottdruleDTForallAppLabel} is such a rule: the instantiation of the
  quantifying variables is not known a priori. The algorithm we present in
  \cref{sec:algorithm} delays this instantiation until more information about it
  (in particular, the set of typing constraints) is collected.

@@

The type polarization (\cref{fig:polarization}) is more involved, as there are
multiple ways to polarize a type: for instance, a variable $[[α]]$ can be
polarized to $[[α⁺]]$ or $[[α⁻]]$. The choice of polarization affects the
execution strategy of the program. 
In particular, a functional type can be represented in either positive (thunked, call-by-value) way:
$[[↓(iP → ↑iQ)]]$ or negative (call-by-name): $[[↓iN → iM]]$.
We chose the positive polarization: every \systemf type is translated to a
\emph{positive} type of \fexists. This choice makes it smoother to lift the
translation to the term level, which will be discussed next.

@@

The term-level translations between the systems are more complex. They are
defined as \emph{elaborations} of the typing derivations. In other words, the
input of the translation is not a term, but a whole typing derivation in the
corresponding system. For example, the subtyping elaboration $[[Γ ⊢ iN ≤ iM ⤳
ft]]$ constructs a \systemf function $[[ft]]$ of type $[[|iN| → |iM|]]$ that
witnesses the given \fexists-subtyping $[[Γ ⊢ iN ≤ iM]]$. A typing elaboration
$[[Γ; Φ ⊢ v : iP ⤳ ft]]$ builds a term $[[ft]]$---a \systemf counterpart of
$[[v]]$---of type $[[|iP|]]$.
